---
title: "Emory Covid-19 Case Study Script"
author: "EpiRHandbook Team"
date: "13 August 2021"
output: 
  officedown::rdocx_document: 
            reference_docx: "template_style.docx"
            keep_md: TRUE
params: 
     surveillance_date: 2021-06-30
editor_options: 
  chunk_output_type: console
---


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
This comment will not show up when you knit the document.
A comment with a title with slashes indicates a name of a code chunk.
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->


```{r setup, include=FALSE}

# hide all code chunks in the output, but show errors 
     # nb. warnings and messages still shown
knitr::opts_chunk$set(echo = FALSE,  # hide all code chunks in output
                      error = TRUE,  # show errors if they appear, but don't stop (produce the word doc)
                      warning = FALSE, # do not show warnings in the output word doc 
                      message = FALSE, # do not show  messages in the output word doc
                      fig.width = 7,         # Figure width
                      fig.height = 6,        # Figure height
                      fig.topcaption = TRUE  # show figure titles on top of plot
                     )

# Ensures the package "pacman" is installed
if (!require("pacman")) {
     install.packages("pacman") }

# install (if necessary) from CRAN and load packages to be used
pacman::p_load(
  officedown, # format MS word document output
  officer,    # add table of contents to output
  rio,        # importing data  
  here,       # relative file pathways 
  skimr,      # get overview of data
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  epikit,     # age_categories() function
  flextable,  # converting tables to pretty images
  sf,         # manage spatial data using a Simple Feature format
  scales,     # define colour schemes for flextables 
  gtsummary,  # summary statistics, tests and regressions 
  labelled,   # create variable labels to be displayed in table outputs
  ggspatial,  # basemaps and scalebars 
  patchwork,  # combining multiple ggplots 
  apyramid,   # plotting age pyramids 
  tidyverse  # data management and visualization

)


## load packages from github
pacman::p_load_gh(
     "R4IDSR/epichecks"   # bivariate colour maps 
)

```


\pagebreak 
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// table_of_contents \\\
--------------------------------------------------------------------------------
This chunk adds a table of contents to your report using the {officedown} package. 
You can also add in a list of tables and figures. 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r table_of_contents}

# table of contents based on headings 
block_toc() 

# list of tables 
block_toc(style = "Table Caption")

# list of figures 
block_toc(style = "Image Caption")

```

\pagebreak 


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// define_reporting_periods \\\
--------------------------------------------------------------------------------
Here we use the date you want to report on, defined in parameters (params) at
top of script to create a date object for filtering datasets, plots/tables, etc. 

We also create a week object to make grouping data easier. In this scenario we 
have defined the week to start on Wednesdays as this is when Fulton County 
releases their reports (but you can choose any other day of the week too). 

From these we are then able to define periods 14 and 28 days prior to our 
report date. We can also make the appropriate labels to auto-populate
table titles. 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r define_reporting_periods}

# create a date object from the parameter
# Minus 7 days surveillance date to account for lag in reporting lab results
surveillance_date <- as.Date(params$surveillance_date) - 7

# create an epiweek object from the date 
# floor_date() rounds down to the closest week here
surveillance_week <- floor_date(surveillance_date,
                          # round by weeks
                          unit = "week", 
                          # define week to start on Wednesday
                          week_start = 3)

# define recent (past 14 days) and previous (28 to 14 days prior)
recent_period   <- seq(surveillance_week  - 13, surveillance_week, by = 1)
previous_period <- seq(surveillance_week  - 27, surveillance_week - 14, by = 1)

# define a text label of date range for the recent period (for table headers)
recent_period_labels <- str_c(
  format(min(recent_period), format = "%m/%d"), 
  "-", 
  format(max(recent_period), format = "%m/%d")
)

# define text label of date range for previous period (for table headers) 
previous_period_labels <- str_c(
  format(min(previous_period), format = "%m/%d"), 
  "-", 
  format(max(previous_period), format = "%m/%d")
)


# define a label for past 28 days (for table captions)
full_period_labels <- str_c(
  format(min(previous_period), format = "%B %d"), 
  "-", 
  format(surveillance_week, format = "%B %d, %Y")
)

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// import \\\
--------------------------------------------------------------------------------

We use the {rio} package for importing our example data - it is very versatile 
and can read in most file types. 

We use the {here} package for defining the path to our file. This is important 
for sharing your script with others (by email or on Sharepoint) - if you used an
"absolute" path, they would need to update the script to match their computer. 

This way your whole R-project folder can be zipped up and moved somewhere else. 

For more details see: 
https://epirhandbook.com/import-and-export.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r import}
  
# import the raw case data set
  # define the path using {here} then pass that to the {rio} import function
  # specify the sheet to read using which (default is to read first sheet)
linelist_raw <- rio::import(
  file = here::here("data", "covid_example_data", "covid_example_data.xlsx"),
  which = "in"
)

# import shapefile
  # for mapping and for extracting population counts for zipcodes
shapefile <- read_sf(
  here::here("data", "covid_example_data", "covid_shapefile", "FultonCountyZipCodes.shp")
)

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// browse_data \\\
--------------------------------------------------------------------------------

Here we take a look at the raw data to get a feel for what needs cleaning. 

We first use the in-built browser with the {base} function View(). 

Then we can use the {base} function summary(), but probably the most comprehensive 
overview is with the {skimr} function. 
You can also view distinct values for variables using the {base} unique() function. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#review
https://epirhandbook.com/descriptive-tables.html#browse-data
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r browse_data, eval = FALSE}

# view your whole dataset interactively (in an excel style format)
View(linelist_raw)

# get summary: 
# mean, median and max values of numeric variables
# counts for categorical variables
# also gives number of NAs
summary(linelist_raw)

# get information about each variable in a dataset 
# nb. “POSIXct” is a type of raw date class 
skim(linelist_raw)

# view unique values contained in variables - useful for categorical variables
# you can run this for any column -- just replace the column name
unique(linelist_raw$case_gender) 


# we can also loop through all categorical variables to look at all possibilities
# see later in the script for introductions to using iteration
purrr::map(
  # all column names (inputs)
  .x = linelist_raw %>% 
    select(where(is_character)) %>% 
    names(), 
  # applies select() then unique() to each column
  .f = ~select(linelist_raw, .x) %>% 
    unique()
  )
```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_colnames \\\
--------------------------------------------------------------------------------

Here we are going to clean the column names of our data set - and store as a new
dataset called "linelist". 

It is possible to use the {janitor} package for automated cleaning of variable 
names - but as there are only a few variables that we want to rename, 
here we will demonstrate using {dplyr} select() function for manually renaming.

Select() can be used either to retain specific columns or to rename them by using
the syntax New name = Old name. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#column-names
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_colnames}

# create a new object called linelist and assign linelist_raw with renamed columns
linelist <- linelist_raw %>% 
  # use select() to retain columns and rename them 
     # NEW name = OLD name
     # aligned for readability
  select( 
    pid                 = PID,
    date_report         = reprt_creationdt_FALSE,      
    date_dob            = case_dob_FALSE,              
    age                 = case_age,                    
    gender              = case_gender,
    race                = case_race,
    eth                 = case_eth,
    zip                 = case_zip,
    # county              = case_county,
    # district            = case_district,
    # state               = case_state,
    contact_id          = Contact_id, 
    date_onset          = sym_startdt_FALSE,
    sym_fever,
    sym_subjfever,
    sym_myalgia,
    sym_losstastesmell,
    sym_sorethroat,
    sym_cough,
    sym_headache,
    sym_resolved,
    date_recovery       = sym_resolveddt_FALSE, 
    contact_hh          = contact_household,
    hospitalized,  
    date_hospitalized   = hosp_admidt_FALSE,
    date_discharge      = hosp_dischdt_FALSE,
    died,  
    died_covid,  
    date_died           = died_dt_FALSE,
    confirmed_case, 
    covid_dx, 
    date_positive       = pos_sampledt_FALSE,
    lat                 = latitude_JITT,
    lon                 = longitude_JITT
    )

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_dates \\\
--------------------------------------------------------------------------------

Here we are going to clean the date variables 

The {base} way to convert columns to Date class is with as.Date(). If the
current format is YYYY-MM-DD or YYYY/MM/DD then no other arguments are needed. 
If the format is different, specify it to format= (for more detail see: 
https://epirhandbook.com/working-with-dates.html#convert-to-date )

Below we also introduce using across() from {dplyr}, which allows you to apply a
function across multiple specified columns. See
https://epirhandbook.com/cleaning-data-and-core-functions.html#clean_across


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_dates}

linelist <- linelist %>% 
  # convert all date columns to date type 
  # note that dates must be formatted correctly - in some cases date parsers from lubridate can be used here
  mutate(
    date_report =       as.Date(date_report),
    date_dob =          as.Date(date_dob),
    date_onset =        as.Date(date_onset)
    # date_recovery =     as.Date(date_recovery),
    # date_hospitalized = as.Date(date_hospitalized),
    # date_discharge =    as.Date(date_discharge),
    # date_died =         as.Date(date_died),
    # date_positive =     as.Date(date_positive)
  ) %>% 
  
  # we can do this in a much faster way by by using across()
  # the apply the as.Date() function to every column with "date" in the name
  # contains() is a tidyselect function (see the handbook for details)
  # the top mutate can be deleted now!
  mutate(across(
    .cols = contains("date"),
    .fns = ~as.Date(.x)
  )) %>%
  
  # remove onset dates prior to 2020
  mutate(across(
    .cols = c(date_report, date_onset, date_hospitalized, date_discharge, date_died),
    .fns  = ~replace(.x, .x < as.Date("2020-01-01"), NA)
    )) %>% 

  # remove dates after the surveillance_date (for this report) from all date columns
  mutate(across(
    .cols = contains("date"),
    .fns  =  ~replace(.x, .x > surveillance_date, NA)
    )) %>%
     
  # create an "epiweek" column from the report date 
  # floor_date() rounds down to the closest week
  mutate(epiweek = floor_date(date_report,
                          # round by weeks
                          unit = "week", 
                          # define week to start on Wednesday
                          week_start = 3)
  )

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_numerics \\\
--------------------------------------------------------------------------------

Here we are going to clean all numeric variables, as well as create some new 
variables based on difference between dates. 

First we will ensure that age is numeric and then fix those with incorrectly 
entered dates (notice that one individual was -20 years old). 

Then we show how to create new numeric variables for the number of days between 
two dates. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#num_cats
https://epirhandbook.com/cleaning-data-and-core-functions.html#clean_case_when

Note the use of if_else() from {dplyr}, which is faster than {base}'s ifelse()
and handles dates better.  
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_numerics}

# Age 
############

linelist <- linelist %>%
  mutate(
    # ensure that age is a numeric variable
    age = as.numeric(age),
    # set those with negative ages and missing DOB to missing 
    # otherwise just leave the age value as is
          # nb. NA_real_ just ensures the variable class is not changed
    age = if_else(age < 0 & is.na(date_dob), NA_real_, age)
  )

     
# Calculating time differences 
##############################

linelist <- linelist %>%
     
  # delay from onset to hospitalization
  mutate(
    # calculate time differences
    days_onset_hosp = as.numeric(date_hospitalized - date_onset),
    # set those under 0 or over 30 to missing
    days_onset_hosp = replace(days_onset_hosp, days_onset_hosp < 0, NA),
    days_onset_hosp = replace(days_onset_hosp, days_onset_hosp > 30, NA)
  ) %>%
     
  # length of hospitalization
  mutate(
    # create outcome date based on whether died or was discharged
    date_outcome = coalesce(date_died, date_discharge),
    # calculate time difference
    days_hosp = as.numeric(date_outcome - date_hospitalized),
    # set those under 0 or over 60 to missing
    days_hosp = replace(days_hosp, days_hosp < 0, NA),
    days_hosp = replace(days_hosp, days_hosp > 60, NA)
  )

## it's a good habit to make sure the variables you're generating seem realistic
# (are you calculating them correctly?)
# summary(linelist$days_hosp)

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_categories \\\
--------------------------------------------------------------------------------

create age groups based on the age variable using 
the age_categories() function from {epikit}. It is also possible to create 
age groups using the {dplyr} case_when() function - but is more involved. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#column-creation-and-transformation
https://epirhandbook.com/cleaning-data-and-core-functions.html#re-code-values
https://epirhandbook.com/characters-and-strings.html?q=regex#regular-expressions-regex
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_categories}

linelist <- linelist %>% 
     
     # create age group variable
     mutate(
       age_group = age_categories(age,
        # define break points
        c(0, 10, 20, 30, 40, 50, 60, 70),
        # whether last break should be highest category
        ceiling = FALSE
     )) %>% 
     
     # recode one value and leave the rest as they are 
     mutate(
       died_covid = if_else(died_covid == "Under Review",
                            "Unknown", died_covid), 
       confirmed_case = if_else(confirmed_case == "Pending", 
                                "Unknown", confirmed_case), 
     
        # force categorical variables to use consistent cases (this can be done for others) 
        sym_myalgia = str_to_title(sym_myalgia),
      ) %>% 
     
     #replace one value and leave the rest, across multiple variables
      mutate(across(
       .cols = c(contact_hh, contains("sym_")),
       .fns  = ~if_else(.x == "Unk", "Unknown", .x)
     )) %>% 
     
     # create a composite category from race and ethnicitiy  
          #  (nb. this sets those that are non-specified in eth to non-hispanic)
     mutate(eth_race = case_when(
       # we evaluate if ethnicity is HISPANIC/LATINO **FIRST**
       # case_when() evaluates in order (if, else if, else if, ... else)
          eth  == "HISPANIC/LATINO"                           ~ "Hispanic, all races", 
          race == "ASIAN"                                     ~ "Asian, NH", 
          race == "BLACK"                                     ~ "Black, NH",
          race == "WHITE"                                     ~ "White, NH",
      # find all instances of NATIVE (covers AMERICAN INDIAN/ALASKA NATIVE **AND** NATIVE HAWAIIAN/PACIFIC ISLANDER)
          str_detect(race, "NATIVE")                          ~ "Other, NH",
          race == "OTHER"                                     ~ "Other, NH", 
          TRUE                                                ~ "Unknown"
     )) %>% 
     
     ## change from upper-case to lower case (with leading capital)
     # mutate(across(
     #   .cols = c(county, state),
     #   .fns = str_to_title
     # )) %>% 
     # 
     
     # recode with searching for string patterns
     # (from "No-Asymptomatic"/"Yes-Symptomatic"/"Unknown")
     mutate(contact_id = case_when(
       str_detect(contact_id, "Yes")     ~ "Yes", 
       str_detect(contact_id, "No")      ~ "No", 
       str_detect(contact_id, "Unknown") ~ "Unknown", 
       TRUE                              ~ "Unknown"
     )) %>% 
  
     # # this section does the same as above, but takes advantage of the way the data is formatted
     # # data is formatted as "Yes/No-A/Symptomatic"
     # # remove text from a string (everything after the dash) - this uses regular expressions (regex)
     # # see handbook section on regex for understanding how this works
     # mutate(contact_id = str_remove(linelist$contact_id, "-.*")) %>% 

     # recode with searching for string patterns 
     mutate(sym_resolved = case_when(
          str_detect(sym_resolved, "Yes")     ~ "Yes", 
          str_detect(sym_resolved, "No")      ~ "No", 
          str_detect(sym_resolved, "Unknown") ~ "Unknown", 
          TRUE                                ~ "Unknown"
     )) %>% 
     
     
     # create a factor from a default numeric class
     mutate(zip = as_factor(zip)) %>% 
     
  
     # replace missing with "Unknown" where relevant 
     mutate(across(
       .cols = c(gender, race, eth, zip,
                 contact_id, contact_hh, 
                 hospitalized, died, died_covid, confirmed_case,
                 contains("sym_"), age_group),
       .fns  = ~fct_explicit_na(.x, na_level = "Unknown")
     )) %>% 
    
     # set levels of a factor (define order)
     mutate(gender      = fct_relevel(gender, "Female", "Male", "Unknown"), 
            eth_race    = fct_relevel(eth_race, 
                                   "Asian, NH", "Black, NH", "White, NH", 
                                   "Hispanic, all races", "Other, NH", "Unknown")
     ) %>% 
     
     # set levels of all factors that are yes/no/unknown 
     mutate(across(
          .cols = c(contact_id, contact_hh, hospitalized, died, died_covid,
                    confirmed_case, contains("sym_")), 
          .fns = ~fct_relevel(.x, "Yes", "No", "Unknown")
     ))
```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// vector_vars \\\
--------------------------------------------------------------------------------

Here we are going to store a list of variable names that we can then use for 
iterating later on rather than typing them out each time. 
In this example we are going to take all the symptom variables and store them in 
a vector. 
It is also possible to simply type these out with quotation marks within a c(...)
argument. 

For more details see: 
https://epirhandbook.com/iteration-loops-and-lists.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r vector_vars}

# define a list of variables for looping over later
symptom_vars <- linelist %>% 
     # choose all columns that contain "sym_" in the name but exclude "sym_resolved"
     select(c(contains("sym_"), -sym_resolved)) %>% 
     # pull the names out 
     names()

# type out the variables of interest
demographic_vars <- c("gender", "age_group", "eth_race")

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// duplicated \\\
--------------------------------------------------------------------------------

Here we remove the duplicates based on having the same pid, gender and date of 
birth. 
Note that this might exclude those which are legitimately reported twice - i.e. 
those who recovered and were reinfected. To deal with these you could create a 
composite variable of the identifiers of interest, flag the duplicates there and 
then add an additional argument for having a report date within six months. 

For more details see: 
https://epirhandbook.com/de-duplication.html 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r duplicated}

# get a data frame of all the duplicates 
     # this is mostly to inspect manually, but can be used for analysing those dropped
duplicates <- linelist %>% 
     get_dupes(pid, gender, date_dob)

linelist <- linelist %>% 
  ## find duplicates based on unique ID, gender and date of birth 
  ## only keep the first occurrence 
  distinct(pid, gender, date_dob, .keep_all = TRUE)

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// filter \\\
--------------------------------------------------------------------------------

Here we are going to filter our data set to only keep relevant cases for analysis. 
We keep those that are reported before our surveillance cut-off date and those 
that are entered as a confirmed case. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#filter-rows
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r filter}

# store those which do not meet our filter criteria 
dropped <- linelist %>% 
     filter(confirmed_case != "Yes" |
              date_report > surveillance_date & 
                !is.na(date_report))


# drop the cases that dont meet the criteria 
linelist <- linelist %>% 
     filter(confirmed_case == "Yes" & 
              date_report <= surveillance_date & 
                 !is.na(date_report))
```


# Descriptive analysis 

This section will be analysing data by time, place, and person to produce descriptive 
tables. 

## Summary 

- As of `r format(surveillance_date, format = "%B %d %Y")`, Fulton County has 
recorded `r nrow(linelist)` confirmed cases of COVID-19.  
- As of `r format(surveillance_date, format = "%B %d %Y")`, Fulton County has 
recorded `r fmt_count(linelist, died == "Yes")` deaths.  
- Among all confirmed cases of COVID-19 in Fulton County, `r fmt_count(linelist, hospitalized == "Yes")`
required hospitalization. 

\pagebreak 

## Time 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// epiweek_table \\\
--------------------------------------------------------------------------------

Here we use the {janitor} tabyl() function to get case counts by calendar week 
(as well as the percentage these contribute to the overall). 

We then create use {flextable} to produce a clean output table. 

For more details see: 
https://epirhandbook.com/descriptive-tables.html
https://epirhandbook.com/tables-for-presentation.html
https://epirhandbook.com/descriptive-tables.html#tbl_janitor
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r epiweek_table, tab.cap="COVID-19 case counts by calendar week in Fulton County", tab.id="timetab"}

# save a quick descriptive table of number of cases reported by week
epiweek_table <- linelist %>% 
  # get counts and percentages 
  tabyl(epiweek) %>% 
  # add the overall counts as a row
  adorn_totals() %>%  
  # change from proportions to percentages (do not add a % sign)
  adorn_pct_formatting(affix_sign = FALSE)
```

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Note that tab.caps cannot have any special characters in them (e.g. "_") otherwise
referencing as below does not work. 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// epicurve \\\
--------------------------------------------------------------------------------

Here we demonstrate how to plot epicurves using the {incidence2} package and edit
it with {ggplot2]. 

For more details see: 
https://epirhandbook.com/epidemic-curves.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r epicurve, fig.cap = "COVID-19 cases by week and hospitalisation status in Fulton County", fig.id = "epicurve"}

# create a sequence of dates for the x axis label
covid_weeks <- seq.Date(
     from = as_date(floor_date(min(linelist$date_onset, na.rm = TRUE), week_start = 3)),
     to = as_date(ceiling_date(max(linelist$date_onset, na.rm = TRUE), week_start = 3)),
     by = "month"
    
)


ggplot(
     data = linelist,
     mapping = aes(
          x = date_onset,
          fill = hospitalized
     )) + 
     
     geom_histogram() + 
     
     scale_x_date(
          breaks = covid_weeks,
          labels = label_date_short()
     ) +
     
     labs(
          x = "",
          y = "Weekly number of cases",
          fill = "Hospitalised",
          caption = paste0("Data as of ", format(surveillance_date, "%d %b %Y"), ". ",
                           fmt_count(linelist, is.na(date_onset))," with missing date of onset")
          
     ) + 
     theme_bw()

```


As you can see from above in Table \@ref(tab:timetab) we can see clear trends in reporting patterns - cases increase in waves and peak in `r filter(epiweek_table, epiweek != "Total") %>% summarise(which.max(n)) %>% pull() %>% epiweek_table[., "epiweek"] %>% as.Date() %>% format(., format = "%B %Y")`. 




\pagebreak 

## Place 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// zip_counts \\\
--------------------------------------------------------------------------------

Here we use {dplyr} functions to get case counts based on ZIP Code and the 
time periods created in the define_reporting_periods code chunk above. 
We store as an object called zip_counts so that it can be merged with population
in the chunks below. 

For more details see: 
https://epirhandbook.com/descriptive-tables.html#dplyr-package
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r zip_counts}

zip_counts <- linelist %>% 
  group_by(zip) %>% 
  # count cases in the appropriate period 
  summarise(
    recent   = sum(date_report %in% recent_period),
    previous = sum(date_report %in% previous_period)
  ) %>% 
  adorn_totals() %>% 
  # a percentage change column and round the digits
  mutate(
    perc_change = round((recent - previous) / previous * 100, digits = 1)
    )

```

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// zip_join \\\
--------------------------------------------------------------------------------

Here, use {dplyr} functions to extract the population counts from our shapefile, 
and then merge these with our zip_counts from above. 
N.b. the left_join() functions preserves all the rows from the first dataset 
provided and only merges the rows from the second that match. 
We then calculate incidence per 10,000 population. 

For more details see: 
https://epirhandbook.com/descriptive-tables.html#dplyr-package
https://epirhandbook.com/joining-data.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r zip_join}

# extract population counts for each zip from the shapefile
zip_pop <- shapefile %>% 
  # change to tibble (otherwise geo-data gets pulled with)
  as_tibble() %>% 
  # only keep zip code and population counts
  select(ZipCode, Population) %>% 
  # add a row with overall counts
  adorn_totals()
  
# merge case counts and population counts
# zip (or ZipCode in the shapefile) variable is the unique identifier
zip_counts <- left_join(zip_counts, 
                        zip_pop, 
                        by = c("zip" = "ZipCode")
                        ) %>% 
  # calculate the incidence 
  mutate(across(
      # for each period (recent and previous)
      .cols = c(recent, previous), 
      # divide each variable by population (and round the outcome)
      .fns = ~round(.x / Population * 10000, digits = 1), 
      # for each period create a new variable with _inc on the end
      .names = "{.col}_inc"), 
    
    # replace NAs in incidence with 0
    across(
      .cols = contains("inc"),
      .fns = ~replace_na(.x, 0)),
    
    perc_change = case_when(
      # fix the outliers: set missing to 0 and infinity (divided by 0) to 100
      is.na(perc_change)       ~ 0,
      is.infinite(perc_change) ~ 100, 
      TRUE                     ~ perc_change
    ))
```

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// zip_table \\\
--------------------------------------------------------------------------------

Here we use {flextable} to produce a publication-ready table. We also demonstrate
how to colour cells based on their values, defining the cut-offs in advance using
{dplyr} case_when(). Colours are called here using HEX-codes but can also be 
referred to by name. 

For more details see: 
https://epirhandbook.com/tables-for-presentation.html
https://colorbrewer2.org/#type=sequential&scheme=Reds&n=3
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r zip_table, tab.cap = "COVID-19 new case counts by ZIP-code", tab.id = "ziptab"}


# pick colours (uncomment next to lines)
# RColorBrewer::brewer.pal(3, "RdYlGn") %>% 
#   scales::show_col()

# choose colours to fill in cells  
row_colour <- case_when(
  # those less than zero will be green (decreasing cases)
  zip_counts$perc_change < 0 ~ "#91CF60", 
  # over zero red (increasing)
  zip_counts$perc_change > 0 ~ "#FC8D59", 
  # missing or zero orange
  TRUE                       ~ "#FFFFBF")


zip_counts %>% 
  # keep the columns of interest and define order
  select(zip, recent, recent_inc, previous, previous_inc, perc_change) %>% 
  # initiate {flextable} to produce styled output table
  flextable() %>% 
  # fill in cells - choose the column and then pass our colour-scheme defined above
  bg(j = "perc_change", 
     bg = row_colour
     ) %>% 
  # add in a header for labeling counts and incidence by period 
    # note the empty columns ("") to fit to the original table headers
  add_header_row(
    values = c("", 
               str_c("Recent 14-day reporting period\n", recent_period_labels), 
               "", 
               str_c("Previous 14-day reporting period\n", previous_period_labels), 
               "", 
               "Change between reporting periods"
               )) %>% 
  # redefine column names based on original names
    # note the different syntax to dplyr::select, here it is old_name = new_name
  set_header_labels(
    zip          = "Zip Code", 
    recent       = "n", 
    recent_inc   = "Incidence", 
    previous     = "n", 
    previous_inc = "Incidence", 
    perc_change  = "%"
  ) %>% 
  # combine the headers cells for the appropriate periods 
  # (i defines rows, j defines columns)
  merge_at(i = 1, j = 2:3, part = "header") %>% 
  merge_at(i = 1, j = 4:5, part = "header") %>% 
  # move the header text to the centre
  align(align = "center", part = "header") %>% 
  # make header text bold 
  bold(part = "header") %>% 
  # make the row with totals in it bold (i.e. the last row in the dataframe)
  bold(i = nrow(zip_counts), part = "body") %>% 
  # add in footnotes for variables (referencing the header cells)
  footnote(j = c(3, 5), part = "header", ref_symbols = c("a"),
           value = as_paragraph("Incidence calculated as cases per 10,000 population by zip code")) %>% 
  footnote(j = 6, part = "header", ref_symbols = c("b"),
           value = as_paragraph("These reflect the percentage increase or decrease of new diagnoses 
                                between the 14 days preceding the past 7 days and the 14 days
                                preceding that.")) %>% 
  # make your table fit to the maximum width of the word document
  set_table_properties(layout = "autofit")

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The tag below will change the page to landscape in the word document output. 
This is part of the {officer} package - and is accompanied by a second tag below
to stop the landscape orientation. 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

<!---BLOCK_LANDSCAPE_START--->

## Person 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// demographics_tab \\\
--------------------------------------------------------------------------------

Here we use {purrr} to iterate over demographics variables (defined in the 
vector_vars code chunk) - to produce tables of counts and percentages for cases
and deaths. 

We use {dplyr} bind_rows() and bind_cols() to combine the various smaller dataframes
of counts in to one large table. 

And then we colour in our {flextable} with different criteria for each of the 
demographic variables of interest. 

For more details see: 
https://epirhandbook.com/tables-for-presentation.html
https://epirhandbook.com/iteration-loops-and-lists.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r demographics_tab, tab.cap = str_c("Cumulative and recent confirmed COVID-19 case and death counts by gender, age, and race/ethnicity in Fulton County, Georgia. Past 28 day period refers to ", full_period_labels), tab.id = "demogtab"}

# get counts tables for measures of interest 
############################################

# we generate 3 summary tables and bind them together
# summary demographic table for gender
dem_gender <- linelist %>% 
  tabyl(gender) %>% 
  select(Characteristic = gender, n, percent)

# summary demographic table for age
dem_age <- linelist %>% 
  tabyl(age_group) %>% 
  select(Characteristic = age_group, n, percent)

# summary demographic table for ethnicity and race
dem_eth_race <- linelist %>% 
  tabyl(eth_race) %>% 
  select(Characteristic = eth_race, n, percent)

# bind all tables together
total_cases <- bind_rows(list(dem_gender, dem_age, dem_eth_race))


# the above can also be done iteratively (we are repeating code)
# this uses the {purrr} package to iterate over each variable
# counts of total cases 
total_cases <- purrr::map(
  # for each variable listed
  .x = demographic_vars,
  # create a table  
  .f = ~tabyl(linelist, .x) %>%
        # only keep variables of interest and rename the variable column 
        # this is so that they are all the same for row binding 
        select("Characteristic" = .x,
               n_cases_total = n,
               perc_cases_total = percent)
  ) %>% 
  # combine rows in to one dataframe
  bind_rows()


# counts of new cases (last 28 days) 
recent_cases <- purrr::map(
  # for each variable listed
  .x = demographic_vars, 
  # filter the linelist for dates on or after 28 days ago
  .f = ~filter(linelist, 
          date_report >= (surveillance_date - 28)) %>% 
        # get counts based on filtered data
        tabyl(.x) %>% 
        # nb we dont keep the characteristic column because it would be duplicated
        select(n_cases_recent = n,
               perc_cases_recent = percent)
  ) %>%
  bind_rows()

# counts of total deaths 
total_deaths <- purrr::map(
  # for each variable listed
  .x = demographic_vars, 
  # filter for those who died 
  .f = ~filter(linelist, 
          died_covid == "Yes") %>% 
        # get counts based on filtered data 
        tabyl(.x, show_na = TRUE) %>%
        select(n_deaths_total = n, perc_deaths_total = percent)
  ) %>% 
  bind_rows()

# counts of new deaths (last 28 days)
recent_deaths <- purrr::map(
  # for each variable listed
  .x = demographic_vars, 
  # filter to those who died in the last 28 days
  .f = ~filter(linelist, 
          died_covid == "Yes" & 
          date_died >= (surveillance_date - 28)) %>% 
        # get counts based on filtered data
        tabyl(.x) %>% 
        select(n_deaths_recent = n, perc_deaths_recent = percent) %>% 
        # add in a variable column (used for colouring later) 
        mutate(variable = .x)
  ) %>% 
  bind_rows()


# total counts for all of the above measures (not by demographic)
overall <- linelist %>% 
  summarise(
    # add in row label 
    Characteristic = "Total",
    # counts of total cases 
    n_cases_total = n(),
    # leave all percentages empty (would just be 100)
    perc_cases_total  = NA, 
    # counts of new cases (last 28 days) 
    n_cases_recent = sum(date_report >= (surveillance_date - 28)), 
    perc_cases_recent  = NA, 
    # counts of total deaths 
    n_deaths_total = sum(died_covid == "Yes"), 
    perc_deaths_total = NA, 
    # counts of new deaths (last 28 days)
    n_deaths_recent = sum(died_covid == "Yes" & 
                          date_died >= (surveillance_date - 28)),
    perc_deaths_recent = NA, 
    # add in a variable column (used for colouring later) 
    variable = "Overall"
  )



# merge tables together 
#######################

# combine all the demographic tables - side by side
demographics_counts <- bind_cols(total_cases, recent_cases, total_deaths, recent_deaths) %>% 
  # mutate each of the proportion columns to be percentages
  mutate(across(
    .cols = contains("perc"),
    .fns = ~round(.x * 100, digits = 1)
    )) 

# add in the totals row at the top of the merged demographics table
demographics_counts <- bind_rows(overall, demographics_counts)



# define colour scheme 
######################

# get the column numbers that are percentages (based on the name) 
percentage_cols <- names(demographics_counts) %>% 
  str_detect("perc") %>% 
  which()

# define colour cut-offs for gender column 
gender_colours <- scales::col_bin(
  # choose colours 
  palette = c("#91CF60", "#FC8D59"), 
  # choose min and max (range)
  domain  = c(0, 100),
  # choose how to split (in this case above and below 50)
  bins    = 2
)

# define colour cut-offs for age column 
age_colours <- scales::col_bin(
  # choose colours
  palette = c("#91CF60","#FFFFBF", "#FC8D59"),
  # choose min and max (range)
  domain  = c(0, 100), 
  # choose cut-off categories 
  bins    = c(0, 5, 20, 100)
)

# define colour cut-offs for ethnicity column 
eth_colours <- scales::col_bin(
  palette = c("#91CF60","#FFFFBF", "#FC8D59"),
  domain  = c(0, 100), 
  bins    = c(0, 10, 40, 100)
)


# create styled table  
######################

demographics_counts %>%
  # initiate flextable to produce styled output table
  flextable(
    # retain variable column for formatting but do not display it
    col_keys = names(demographics_counts)[-10]
  ) %>%
  # redefine column names based on original names
  set_header_labels(
    "n_cases_total"       = "Total Confirmed Cases",
    "perc_cases_total" = "% of Total Cases",
    "n_cases_recent"       = "Confirmed Cases past 28 days",
    "perc_cases_recent" = "% of Confirmed Cases past 28 days",
    "n_deaths_total"       = "Total Confirmed Deaths",
    "perc_deaths_total" = "% of Total Deaths",
    "n_deaths_recent"       = "Confirmed Deaths past 28 days",
    "perc_deaths_recent" = "% of Confirmed Deaths past 28 days"
  ) %>%
  # move the header text to the centre
  align(align = "center", part = "header") %>%
  # make header text bold
  bold(part = "header") %>%
  # make the totals row bold (i.e. first row)
  bold(i = 1, part = "body") %>%
  # fill in the cells
  # choose the rows with gender counts
  bg(i = ~variable == "gender",
     # choose the columns with percentages in them
     j = percentage_cols,
     # fill in based on previous defined cut-offs
     bg = gender_colours) %>%
  bg(i = ~variable == "age_group",
     j = percentage_cols, bg = age_colours) %>%
  bg(i = ~variable == "eth_race",
     j = percentage_cols, bg = eth_colours) %>%
  # add horizontal lines after the cells with totals and unknowns
    # (short-cut to find row ending of each demographic variable)
  hline(i = ~Characteristic %in% c("Total", "Unknown")) %>%
  # add in footnotes for rows counting unknowns (reference in first column)
  footnote(i = ~Characteristic == "Unknown", j = 1, part = "body", ref_symbols = c("a"),
           value = as_paragraph("Unknown includes cases not yet interviewed")) %>%
  # add in footnote for deaths counts (ref in the header)
  footnote(i = 1, j = c(6, 8), part = "header", ref_symbols = c("b"),
           value = as_paragraph("Deaths refer to all persons who had a positive PCR test result
                                for Covid-19 and there is evidence that COVID-19 was the cause of
                                death or a significant contributor to their death.")) %>%
  # make your table fit to the maximum width of the word document
  set_table_properties(layout = "autofit")

```

<!---BLOCK_LANDSCAPE_STOP--->

### Risk factors for mortality

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// rf_unknowns \\\
--------------------------------------------------------------------------------

Here we need to define our variables of interest to investigate as risk factors. 
We then need to swap factor levels so that the appropriate levels are selected 
as the exposure level in regression. 
We then need to keep only complete cases (i.e. that don't have any missings
for any of the variables of interest). 
Finally, we  use the {labelled} package to assign text for each variable name to appear 
in output tables. 

For more details see: 
https://epirhandbook.com/univariate-and-multivariable-regression.html#store-explanatory-variables
https://epirhandbook.com/cleaning-data-and-core-functions.html#re-code-values
http://larmarange.github.io/labelled/
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r rf_unknowns}

# define variables of interest (save typing them out later) 
descriptive_vars <- c("gender", 
                      "age_group",
                      "eth_race",
                      symptom_vars,
                      "hospitalized",
                      "days_hosp")

# filter dataset  
rf_data <- linelist %>% 
  # only keep variables of interest
  select(died_covid, age, all_of(descriptive_vars)) %>% 
  # set unknown back to NA for all factor variables
  mutate(across(
    .cols = where(is.factor),
    .fns = ~fct_recode(.x, NULL = "Unknown"))) %>% 
  # flip factor levels (so that the reference values are correct)
  mutate(eth_race = fct_infreq(eth_race)) %>% 
  mutate(gender = fct_relevel(gender, "Female", "Male")) %>% 
  mutate(across(all_of(c("died_covid", symptom_vars, "hospitalized")), 
                ~fct_relevel(.x, "No", "Yes")
                )) %>% 
  # only keep rows with complete data for all variables of interest
  # note that this will drop rows where **ANY** of the listed variables are NA
  drop_na(any_of(c("died_covid", "age", descriptive_vars)))


# define variable labels to show in output tables 
rf_data <- rf_data %>%
  set_variable_labels(
    died_covid = "Died",
    age = "Age (years)",
    gender = "Gender",
    age_group = "Age group (years)",
    eth_race = "Ethnicity",
    sym_fever = "Fever",
    sym_subjfever = "Subjective fever",
    sym_myalgia = "Myalgia",
    sym_losstastesmell = "Loss taste/smell",
    sym_sorethroat = "Sore throat",
    sym_cough = "Cough",
    sym_headache = "Headache",
    hospitalized = "Hospitalized",
    days_hosp = "Days in hospital"
  )

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// simple_stats \\\
--------------------------------------------------------------------------------

Here we demonstrate how to do simple statistical tests using the {gtsummary} 
package. We can define which tests to run for which variables and then modify the 
table format afterwards, also using {gtsummary}. The syntax for formatting is 
similar to {flextable} and as this needs to be in {flextable} for MS word anyway, 
you could also do all the formatting using {flextable}.  

For more details see: 
https://epirhandbook.com/simple-statistical-tests.html#stats_gt
https://epirhandbook.com/tables-for-presentation.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r simple_stats, tab.cap = "Demographic characteristics and association with mortality among COVID-19 cases", tab.id = "statstab"}

rf_data %>%
  # keep variables of interest
  select(died_covid, gender, eth_race, age, days_hosp) %>%
  # produce summary table and specify grouping variable
  tbl_summary(
    by = died_covid
  ) %>%
  # specify what test to perform
  add_p(
    list(
      all_continuous() ~ "kruskal.test",
      eth_race ~ "kruskal.test",
      all_dichotomous() ~ "chisq.test"
    )
  ) %>%
  # edit what the column headers say (using {gtsummary})
  # nb. {n} automatically shows the number in that group and \n is a linebreak
  modify_header(update = list(
    stat_1 ~ "**Dead**\n (N={n})",
    stat_2 ~ "**Alive**\n (N={n})"
  )) %>%
  # edit what it says in the footnote (using {gtsummary})
  modify_footnote(update = list(
    all_stat_cols() ~ "n (%) for categorical;\n median (IQR) for continuous",
    p.value ~ "Pearson's Chi-squared test for dichotomous;\n Kruskal-Wallis rank sum test for continuous and categorical"
  )) %>%
  # change to flextable format
  as_flex_table() %>%
  # make header text bold (using {flextable})
  bold(part = "header")
```
  

\pagebreak 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// regression \\\
--------------------------------------------------------------------------------

Here we demonstrate how to do univariate regression to produce odds ratios with 
{gtsummary}. We first create a regression table, then a counts table and then 
finally merge the two and format the output. 
Note that a cox regression taking into account observation time might be a 
more appropriate analysis given the question on mortality - you could do this 
using {gtsummary} but take a look at the epiRhandbook page on survival analysis.

For more details see: 
https://epirhandbook.com/univariate-and-multivariable-regression.html
https://epirhandbook.com/survival-analysis.html
https://epirhandbook.com/tables-for-presentation.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r regression, tab.cap = "Univariate regression investigating associations with mortality among COVID-19 cases", tab.id = "regrtab"}

# produce table with regression estimates
regress_tab <- rf_data %>%
  # drop variables not interested in 
  select(-age_group) %>%
  # produce univariate table
  tbl_uvregression(
    # define outcome variable
    y = died_covid, 
    # define regression want to run (generalised linear model)
    method = glm, 
    # define what type of glm want to run (logistic)
    method.args = list(family = binomial), 
    # exponentiate to produce odds ratios (rather than log odds)
    exponentiate = TRUE, 
    # do not show the overall counts (this is done in cross_tab below)
    hide_n = TRUE,
    ## uncomment this line if you want to not show reference rows
    # show_single_row = c(symptom_vars, gender, hospitalized),
    ## note: NULL at the end allows you to have a comma before a commented out row
    NULL
  )

# produce table with counts by outcome (using the data fed to the regression above)
cross_tab <- regress_tab$inputs$data %>%
  tbl_summary(
    # group by outcome 
    by = died_covid,
    ## uncomment this line if you only want to show the "Male" row for gender
    ## this would be run if you also uncommented the single_row in regression above
    # value = list(gender ~"Male"),
    ## show all levels (otherwise only shows the "Yes" level)
    type = list(all_dichotomous() ~ "categorical"),
    ## note: NULL at the end allows you to have a comma before a commented out row
    NULL
  )

# combine tables 
tbl_merge(list(cross_tab, regress_tab)) %>%
  # edit what it says in the grouping headers 
  modify_spanning_header(update = list(
    c("stat_1_1","stat_2_1") ~ "Died",
    c("estimate_2", "ci_2", "p.value_2") ~ "Univariate regression")
    ) %>% 
  # edit what it says in the footnote (using {gtsummary})
  modify_footnote(update = list(
    all_stat_cols() ~ "n (%) for categorical;\n median (IQR) for continuous")
    ) %>% 
  # change to flextable format
  as_flex_table() %>%
  # make header text bold (using {flextable})
  bold(part = "header") %>% 
  # make your table fit to the maximum width of the word document
  set_table_properties(layout = "autofit")
```

As seen in Table \@ref(tab:regrtab), mortality was associated with being male 
`r inline_text(regress_tab, variable = gender, level = "Male", pattern = "(OR {estimate}; 95% CI {conf.low}, {conf.high})")`, being white `r inline_text(regress_tab, variable = eth_race, level = "White, NH", pattern = "(OR {estimate}; 95% CI {conf.low}, {conf.high})")`and longer 
hospital stays `r inline_text(regress_tab, variable = days_hosp, pattern = "(OR {estimate}; 95% CI {conf.low}, {conf.high})")`.

\pagebreak 

# Visualisation 


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// ggplot_theme \\\
--------------------------------------------------------------------------------

Here we define a {ggplot2} theme (which is applied to all plots) and then define 
some details for theme which can be applied to plots when needed. 

For more details see: 
https://epirhandbook.com/ggplot-basics.html#ggplot_basics_themes
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r ggplot_theme}

# Set default options for plots and charts
  # set default text size for plots (base_size)
  # give classic black/white axes for plots
ggplot2::theme_set(theme_classic(base_size = 18))

# sets the theme in ggplot for how plots should appear
epicurve_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
  panel.grid.major.x = element_line(color = "grey60", linetype = 3),
  panel.grid.major.y = element_line(color = "grey60", linetype = 3), 
  axis.line = element_line(colour = "black"))
```


## Person 

As we see below in Figure \@ref(fig:agehospplot) there appears to be an 
association between older age, length of hospital stay and mortality. 
We saw this in Table \@ref(tab:regrtab) as well for the individual associations, 
although we did not check for interaction and confounding. 



\pagebreak 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// ethnicity_counts \\\
--------------------------------------------------------------------------------

Here we demonstrate how to plot bar charts with {ggplot2}. 

For more details see: 
https://epirhandbook.com/ggplot-basics.html#geoms
https://epirhandbook.com/ggplot-tips.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r ethnicity_counts, fig.cap = "COVID-19 cases by ethnicity", fig.id = "ethnicityplot"}

# open a plot with the linelist data
ggplot(linelist) +
  # add bars 
  geom_bar(
    mapping = aes(
      # plot the number of cases by ethnicity (ordered in reverse frequency)
      x = fct_rev(fct_infreq(eth_race)),
      # stack bars and colour by died (ordered in reverse frequency)
      fill = fct_rev(fct_infreq(died))
    )
  ) +
  # flip the x and y axes 
  coord_flip() +
  # make the x axes start at the origin (nb axes flipped)
  scale_y_continuous(expand = c(0, 0), 
                     # define where to label xaxis (nb axes flipped )
                     breaks = seq(from = 0,
                                  to = 35000,
                                  by = 5000)) + 
  # add in labels 
  labs(
    # set the axes titles (nb axes flipped)
    x = "Race and Ethnicity",
    y = "Cases (n)",
    caption = "Fictional COVID-19 data",
    fill = "Deceased"
    ) + 
  # apply the theme we defined in ggplot_theme code chunk above 
  epicurve_theme

```


\pagebreak 

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// age_pyramid \\\
--------------------------------------------------------------------------------

Here we demonstrate how to plot age pyramids using the {apyramid} package and edit
it with {ggplot2]. 

For more details see: 
https://epirhandbook.com/demographic-pyramids-and-likert-scales.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r age_pyramid, fig.cap = "Age and Gender of COVID-19 cases, Fulton County", fig.id = "agepyramid"}

# prepare dataset

# start a new dataframe (as dont want to overwrite the original)
linelist_2g <- linelist %>% 
  # update the gender and age_group columns
  mutate(across(.cols = c(gender, age_group), 
                .fns = ~{
                  # replace "Unknown" with NA
                  .x = na_if(.x, "Unknown") 
                  # drop "Unknown" from the factor levels 
                  .x = fct_drop(.x)
                }))

# plot age pyramid 
age_pyramid(
  data = linelist_2g,
  age_group = "age_group",
  split_by = "gender",
  # Show as percentages of total cases
  proportional = TRUE,
  # remove guide line for mid-point
  show_midpoint = FALSE) +
  # set theme to basic 
  theme_minimal() +
  # add labels 
  labs(
    title = "",
    subtitle = ,
    x = "Age group",
    y = "Percent of total",
    fill = "Gender",
    # use str_glue to set dynamic captions 
    # {missing} is defined in the second argument below
    caption = str_glue(
      "{missing} cases missing either age or gender are not shown. \n Fictional COVID-19 data",
      missing = linelist_2g %>%
        filter(is.na(gender) | is.na(age_group)) %>%
        nrow()
      )
    )

```


\pagebreak 

## Time 



