# Preface {.unnumbered}

This is a Quarto book.

To learn more about Quarto books visit <https://quarto.org/docs/books>.

```{r}
pacman::p_load(
  rio,
  here,
  janitor,
  DT,
  tidyverse
)
```

See [Fulton County Espanol](fulton.qmd#fulton-es)

| Name   | Link                        | Topics         | Geography | Scale | Tool | Language |
|---------|-------------------|-------|-----------|---------|---------|---------|
| Fulton | [Fulton](fulton.qmd#fulton) | Outbreak; GIS; | USA       | Local | R    | English  |
| Oswego | [Oswego](oswego.qmd#oswego) | Outbreak       | USA       | Local | R    | Spanish  |

```{r, eval=F}
table_cases <- tribble(
  ~Name, ~Link, ~Geography, ~Scale, ~Tool, ~Language, ~Topics,
  "Fulton COVID-19", [Fulton](fulton.qmd#fulton), "USA", "Local", "R", "English", "Outbreak reports; GIS; epidemic curves",
  "Oswego ES", "Link", "USA", "Local", "R", "Spanish", "Outbreak response"
)
```

Filterable version of table (but will need to insert weblinks, not references)

```{r, eval=F}
DT::datatable(table_cases, escape = F)
```

```{r}
1 + 1
```
