---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, echo=F}
pacman::p_load(
  rio,
  here,
  janitor,
  ISOcodes,
  tidyverse,
  googlesheets4,
  reactable
)
```

## Welcome to our Open Case Study Repository!

A case study is a fun, interactive way to learn—guiding you through real-life scenarios with tasks and questions. You’ll find a variety of topics, skill levels, and diseases below, most of which take just a few hours to complete. You can filter and order the rows to pick the one that fits you best. Click the link corresponding to your preferred language to open it up. The data used in the case studies is either generated or anonymized. 

For more information on how to use the case studies, click the [How-to
Guide](pages/instructions.html). 

```{r, echo = F, eval=T, warning = F, message= F}

# define current language of interest 
# (this allows us to just copy paste this code chunk to other langs and switch, 
# ideally there would be a way to auto switch using inputs from babelquarto?)
curr_lang <- "en"
curr_lang_iso3 <- "eng"

# Specify the URL of the Google Sheet (which lists case studies)
sheet_url <- "https://docs.google.com/spreadsheets/d/1wrZtM3XjUtbDXD7PVLOHdd-H92hq_-WAQoIpRatjhlc/edit?gid=598142358"


# Read the data from the Google Sheet
table_raw <- read_sheet(sheet_url, sheet = "Publishing")

# Read in the column names translation sheet from Google sheet
translations <- read_sheet(sheet_url, sheet = "TranslationCols")


# filter to show current language 
# still have links to other languages by combining
table <- table_raw %>%
  mutate(casestudy_identifier= if_else(publish=="yes",
                                       casestudy_identifier, 
                                       "under_construction"), 
         # backup language for filter usage 
         og_language = language) %>% 
  # recode language and country 
  mutate(language = matchmaker::match_vec(language, 
                                               dictionary = ISOcodes::ISO_639_3, 
                                               from = "Part1", to = curr_lang_iso3),
         language = sub("\\(.*", "", language), 
         country = toupper(country),
         country = matchmaker::match_vec(country, 
                                         dictionary = ISOcodes::ISO_3166_1, 
                                         from = "Alpha_3", to = "Name")) %>% 
  mutate(
    # change languages to clickable links 
    language = paste0(
      '<a href="pages/', 
      casestudy_identifier, 
      if_else(og_language != "en", paste0(".", og_language), ""),
      '.html" target="_self">', 
      language, 
      '</a>'
    )
  )

# Find available languages using group_identifier to combine languages
find_langs <- table %>% 
  group_by(group_identifier) %>% 
  summarize(language = paste(unique(language), collapse = ", "), 
                   .groups = 'drop')  


# Join with the original data to retain 'name' and handle suffixes
output <- table %>% 
  #filter(og_language == curr_lang) %>% 
  # Restrict columns  
  select(group_identifier, name, creator, tool, country, scale, subject,
                `technical complexity`, `methodological complexity`) %>% 
  unique() %>% 
  left_join(find_langs, by = "group_identifier") %>% 
  # rearrange to have language further forward (not sure why this is not agreeing)
 select(name, creator, language, tool, subject, `technical complexity`, `methodological complexity`)
  

# Create table
reactable(output,
  columns = list(
    name = colDef(name = matchmaker::match_vec("name", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang), 
                  style = list(width = "300px",
                               fontWeight = "bold"),
                  headerStyle = list(width = "300px")),
    creator = colDef(name = matchmaker::match_vec("creator", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang)), 
    language = colDef(name = matchmaker::match_vec("language", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang), 
                      html = TRUE,
                      align = "left"),
    tool = colDef(name = matchmaker::match_vec("tool", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang),
                  style = list(width = "10px"),
                  headerStyle = list(width = "10px")), 
    # country = colDef(name = matchmaker::match_vec("country", 
    #                                            dictionary = translations, 
    #                                            from = "og", to = curr_lang), 
    # scale = colDef(name = matchmaker::match_vec("scale", 
    #                                            dictionary = translations, 
    #                                            from = "og", to = curr_lang), 
    subject = colDef(name = matchmaker::match_vec("subject", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang)),
    `technical complexity` = colDef(name = matchmaker::match_vec("technical complexity", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang),
                                    align = "center",
                                    style = list(width = "140px"),
                                    headerStyle = list(width = "140px")),
    `methodological complexity` = colDef(name = matchmaker::match_vec("methodological complexity", 
                                               dictionary = translations, 
                                               from = "og", to = curr_lang), 
                                         align = "center",
                                         style = list(width = "145px"),
                                         headerStyle = list(width = "145px"))
  ),
  defaultPageSize = 10,
  filterable = TRUE,
  sortable = TRUE,
  style = list(backgroundColor = "white", 
               color = "black", 
               fontFamily = "Arial",  
                fontSize = "17px",
               border = "1px solid #000"), # Blue background with white text),
  rowStyle = list(backgroundColor = "white", 
               color = "black", 
               fontFamily = "Arial",  
                fontSize = "14px"), # Blue background with white text),
)
```


## About us

[**Applied Epi**](https://www.appliedepi.org/) is a nonprofit organisation and grassroots movement of frontline epidemiologists from around the world. Your encouragement and feedback is most welcome:

-   Visit our [**website**](https://www.appliedepi.org/) and [**join our contact list**](https://docs.google.com/forms/d/e/1FAIpQLSdtk7Mg32lYQ3ucbIM-kIwIRxgRwDfG5TJHsVAbFjQ1i6fwiw/viewform?usp=send_form)

-   **contact\@appliedepi.org**, tweet [**\@appliedepi**](https://x.com/appliedepi), or [**LinkedIn**](https://www.linkedin.com/company/appliedepi/?originalSubdomain=it)

-   Submit issues to our **Github repository**