---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Case-studies Open Repository {.unnumbered}

```{r, echo=F}
pacman::p_load(
  rio,
  here,
  janitor,
  DT,
  tidyverse
)
```

**Objective**: This open repository of case studies can be used as a guide on how to implement different epidemiological methods in different situations using a variety of tools

# How to use this repository {.unnumbered}

# Acknowledgements {.unnumbered}

## Authors {.unnumbered}


# List of Case-studies {.unnumbered}


This is an internal link: See [Fulton County English](fulton.qmd#fulton-en)

This is another kind of internal link that makes use of a tag: [Chapter -@sec-fulton-en]

This is a static table to test links:

| Name   | Link                           | Topics         | Geography | Scale | Tool | Language |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Fulton | [Fulton](fulton.qmd#fulton-en) | Outbreak; GIS; | USA       | Local | R    | English  |
| Oswego | [Oswego](oswego.qmd#oswego-es) | Outbreak       | USA       | Local | R    | Spanish  |

```{r, eval=F, echo=F}
table_cases <- tribble(
  ~Name, ~Link, ~Geography, ~Scale, ~Tool, ~Language, ~Topics,
  "Fulton COVID-19", @sec-fulton, "USA", "Local", "R", "English", "Outbreak reports; GIS; epidemic curves",
  "Oswego ES", [Oswego](oswego.qmd#oswego), "USA", "Local", "R", "Spanish", "Outbreak response"
)
```

```{r, echo=F}
table_cases <- tribble(
  ~Name, ~Link, ~Geography, ~Scale, ~Tool, ~Language, ~Topics,
  "Fulton COVID-19", "<a @sec-fulton-en>Fulton</a>", "USA", "Local", "R", "English", "Outbreak reports; GIS; epidemic curves", "Oswego ES", "<a @sec-oswego>Oswego</a>", "USA", "Local", "R", "Spanish", "Outbreak response"
  )

table_cases <- table_cases %>% 
  mutate(across(.cols = everything(),
                .fns = ~as.factor(.x)))

```

Filterable version of table (but will need to insert weblinks, not references as they do not seem to work)

```{r, echo = F, eval = T}

DT::datatable(table_cases, escape = F, filter = "top")

```
