# Landing page {.unnumbered}

This is a Quarto book. To learn more about Quarto books visit <https://quarto.org/docs/books>.

```{r, echo=F}
pacman::p_load(
  rio,
  here,
  janitor,
  DT,
  tidyverse
)
```

This is an internal link: See [Fulton County English](fulton.qmd#fulton-en)

This is another kind of internal link that makes use of a tag: \@sec-fulton-en

This is a static table to test links:

| Name   | Link                        | Topics         | Geography | Scale | Tool | Language |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Fulton | [Fulton](fulton.qmd#fulton-en) | Outbreak; GIS; | USA       | Local | R    | English  |
| Oswego | [Oswego](oswego.qmd#oswego-es) | Outbreak       | USA       | Local | R    | Spanish  |


Can we incorporate a keywords column??
Otherwise have Search bar for whole book, like we do for the Epi R Handbook.


This is an attempt dynamic table made using {DT} - LINKS TO CHAPTERS DO NOT WORK because putting a @ or a @ inside the `tribble()` function seems to break the table. An alternative would be to put the whole website address (e.g. www.appliedepi.org/case/oswego...) but we won't know this till the end.

Putting internal links in DT does not seem to work. Might need to think of other solution...

```{r, eval=F, echo=T}
table_cases <- tribble(
  ~Name, ~Link, ~Geography, ~Scale, ~Tool, ~Language, ~Topics,
  "Fulton COVID-19", @sec-fulton, "USA", "Local", "R", "English", "Outbreak reports; GIS; epidemic curves",
  "Oswego ES", [Oswego](oswego.qmd#oswego), "USA", "Local", "R", "Spanish", "Outbreak response"
)
```

Filterable version of table (but will need to insert weblinks, not references)

```{r, eval=F}
DT::datatable(table_cases, escape = F)
```
