---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, echo=F}
pacman::p_load(
  rio,
  here,
  janitor,
  DT,
  tidyverse
)
```

## Welcome to our Case Study Repository!

Here, you can practice technical and analytical skills using real-life scenarios. The case studies, contributed by FETPs and health ministries worldwide, feature data that is either generated or anonymized for training purposes.

Find a case study of interest with the table below, and click on the link to open it in the appropriate language for you. 

```{r echo = F}
table_cases <- tribble(
  ~Name, ~Link, ~Creator,  ~Language, ~Tool, ~Location, ~Scale, ~Disease, ~Keywords, ~Technical_Complexity, ~Methodological_Complexity, 
  "Fulton", 'pages/fulton-en.html', "Applied Epi", "English", "R", "United States", "Local", "COVID-19", "COVID-19; SARS-COV-2; Outbreak", "Intermediate", "Basic",  
  "Oswego ES", "pages/oswego-es.html", "US CDC (adapted)", "Español", "R", "Estados Unidos", "Local", "Gastrointestinal", "Gastrointestinal;Brote", "Intermedio", "Intermedio"
)

```


```{r, echo = F, eval = T}

# Render the table with DT, making the 'Link' column clickable with custom text
datatable(table_cases, escape = FALSE, filter = "top", options = list(
  columnDefs = list(
    list(targets = 2, render = JS(
      "function(data, type, row, meta) {
        if(type === 'display') {
          var linkText = 'Link'; // Use the 'Name' column as the display text
          data = '<a href=\"' + data + '\">' + linkText + '</a>';
        }
        return data;
      }"
    ))
  )
), callback = JS("table.page.len(5).draw();"))

```

## About us

[**Applied Epi**](https://www.appliedepi.org/) is a nonprofit organisation and grassroots movement of frontline epis from around the world. We write in our spare time to offer this resource to the community. Your encouragement and feedback is most welcome:

-   Visit our [**website**](https://www.appliedepi.org/) and [**join our contact list**](https://docs.google.com/forms/d/e/1FAIpQLSdtk7Mg32lYQ3ucbIM-kIwIRxgRwDfG5TJHsVAbFjQ1i6fwiw/viewform?usp=send_form)

-   **contact\@appliedepi.org**, tweet [**\@appliedepi**](https://x.com/appliedepi), or [**LinkedIn**](https://www.linkedin.com/company/appliedepi/?originalSubdomain=it)

-   Submit issues to our **Github repository**