---
title: "E-pox: Importing, cleaning and describing data"
author: "Xanthi Andrianou, Gianfranco Spiteri (ECDC EI Group)"
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook:
    split_by: none
    self_contained: true
    code_folding: hide
---


```{r E-pox-Rpractical-1, include=FALSE}
# Basic script to install packages

# List of packages to install
packages_to_install <- c("tidyverse", "janitor", "lubridate",
                        "jsonlite", "readxl",
                        "skimr", "gtsummary", "knitr", "apyramid")

# For each package in the list, install if not already installed
for (package in packages_to_install) {
 if (!requireNamespace(package, quietly = TRUE)) {
   install.packages(package, dependencies = TRUE)
 }
}

# Remove the list
rm(packages_to_install)

knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  include = TRUE,
  results = "markup", # "asis", #"hold",
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  error = FALSE, # stop on errors if F
  tidy = "styler", # Use styler::style_text() to reformat the code
  tidy.opts = list(scope = "tokens", strict = TRUE, indent_by = 3),
  # tidy = "formatR", # Use formatR::tidy_source() to reformat the code
  # tidy.opts = list(blank = FALSE, width.cutoff = 60),
  background = "#F7F7F7", # gre97  ## "#e5e5e5" # gre90
  cache = FALSE,
  dev = c("png", "pdf"),
  dev.args = list(png = list(type = "cairo-png")),
  dpi = if (knitr::is_html_output()) 96 else 300,
  fig.width = 7,
  fig.asp = 0.618, ## golden ratio
  out.width = "70%",
  fig.align = "center",
  fig.show = "hold",
  out.width = "100%",
  webshot = "webshot"
)


library(tidyverse)
library(lubridate)
library(janitor)
library(glue)
library(knitr)
```


# Importing

## Case-based data

```{r E-pox-Rpractical-2 }
CB_Data0<-read.csv("data/raw/E_pox_case_based_data.csv")

skimr::skim(CB_Data0)

# Importing another file type (see ppt)
CB_Data0_json<-jsonlite::fromJSON("data/raw/E_pox_case_based_data.json")
CB_Data0_json<-rio::import("data/raw/E_pox_case_based_data.json")

CB_Data0_xlsx<-readxl::read_excel(path = "input_data/E_pox_case_based_data.xlsx")
```

## Aggregated data

```{r E-pox-Rpractical-3 }

Agg_Data0<-read.csv("input_data/E_pox_aggregated_data.csv")

skimr::skim(CB_Data0)

Agg_Data0_json<-jsonlite::fromJSON("input_data/E_pox_aggregated_data.json")


Agg_Data0_xlsx<-readxl::read_excel(path = "input_data/E_pox_aggregated_data.xlsx")

# Remove the files we do not need 
# within rm() we ask for the objects containing 
# the pattern "json" or "xlsx" to be removed from the environment
rm(list = ls(pattern = "json|xlsx"))

```

# Cleaning

## Case-based data

```{r E-pox-Rpractical-4 }
# Start working using the CB_Data0 dataset

str(CB_Data0)
summary(CB_Data0)

head(CB_Data0)
glimpse(CB_Data0)

# Working on the date variables

CB_Data <- CB_Data0 %>% 
  # convert one date variable
  mutate(DateOfNotification=ymd(DateOfNotification)) %>% 
  # Convert the date variables to date
  mutate(across(starts_with("Date"),
                ~ ymd(.x))) %>% 
  # Add a variable with the week
  mutate(WeekDate=floor_date(DateOfNotification, unit="week", week_start="Monday")) %>% 
  mutate(
    Gender = case_when(
      Gender == "F" ~ "Female",
      Gender == "M" ~ "Male",
      Gender == "O" ~ "Other",
      Gender == "UNK" ~ "Unknown",
      .default = Gender
    )
  )


head(CB_Data)
skimr::skim(CB_Data)
glimpse(CB_Data)

plot(CB_Data$DateOfNotification)
hist(CB_Data$DateOfDiagnosis, breaks="weeks")
hist(CB_Data$DateOfOnset, breaks="weeks")

skimr::skim(CB_Data)

# Cross-tabulate the country info and the date of notification
janitor::tabyl(dat = CB_Data, 
               WeekDate, 
               Country)

# Cross-tabulate the country info and the gender variable
janitor::tabyl(dat = CB_Data, 
               Gender, 
               Country)
```



## Aggregated data

```{r E-pox-Rpractical-5 }
# Working on the Agg_Data0 dataset

str(Agg_Data0)
names(Agg_Data0)

summary(Agg_Data0)

Agg_Data <- Agg_Data0 %>% 
  # Convert the date variables to date
  mutate(across(.cols= starts_with("Date"),
                .fns = ~ ymd(.x))) %>% 
  # Convert character columns to factors:
  mutate(across(.cols = where(is.character), 
                .fns = ~ as.factor(.x))) %>% 
  # Add a variable with the week
  mutate(WeekDate=floor_date(DateRep, unit="week", week_start="Monday")) 
  
str(Agg_Data)

summary(Agg_Data)


# nrow() used below gives the total number of rows in the dataset
```

# Basic descriptives of the case-based data

Total number of cases: `r nrow(CB_Data)`

## Summary of the characteristics of the cases (tables and graphs)

### Summary by country

```{r E-pox-Rpractical-6 }

CB_table <- CB_Data %>%
  # Select columns starting with "Date"
  select(Country) %>%
 # Create summary table (attention to handling of missing values)
  gtsummary::tbl_summary(missing = "ifany") 
CB_table

```

### Time

```{r E-pox-Rpractical-7, echo=TRUE}
# Checking the different variables

CB_table1 <- CB_Data %>%
  # Select columns starting with "Date"
  select(starts_with("Date")) %>%
 # Create summary table (attention to handling of missing values)
  gtsummary::tbl_summary(missing = "ifany") 

CB_table1

```

#### Epicurves (overall and by country)

```{r E-pox-Rpractical-8, echo=TRUE, fig.height=7, fig.width=9}
# First we can nake an object to summarise the data we want to
# show on the epicurve
EpiCurve_object<- CB_Data %>% 
  group_by(WeekDate) %>% 
  summarise(n=n(), .groups = "drop") 

head(EpiCurve_object)

EpiCurve <- EpiCurve_object %>% 
  ggplot() +
  geom_col(aes(x = WeekDate, y = n), fill="darkgreen", color="white", alpha=0.8) +
  scale_fill_viridis_d() +
  scale_x_date(breaks = "2 weeks")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        axis.text.x = element_text(angle=90),
        legend.title = element_blank()) +
  labs(title="E-pox cases reported in 2022")

EpiCurve
```


```{r E-pox-Rpractical-9,  echo=FALSE, fig.height=7, fig.width=9}
# The dataset to be used for the epicurve (and in general 
# in ggplot()) can be created right before the plot and it 
# can be passed directly to the ggplot()

EpiCurve_byCountry<-CB_Data %>% 
  group_by(Country, WeekDate) %>% 
  summarise(n=n(), .groups = "drop") %>%  
  ungroup() %>% 
  ggplot(.) +
  geom_col(aes(x = WeekDate, y = n, fill = Country), color="white", alpha=0.8) +
  scale_fill_viridis_d() +
  scale_x_date(breaks = "2 weeks")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        axis.text.x = element_text(angle=90),
        legend.title = element_blank()) +
  labs(title="E-pox cases reported by country in 2022")

EpiCurve_byCountry
```

#### Plot the dates of onset, diagnosis, and notification

In the plots below, we use the dates of onset, diagnosis and notification and put them together to see whether there are difference and visualize the information to check if there are large delays. First, we try with dates and the plot is a messy (not easy to interpret.

```{r E-pox-Rpractical-10, fig.height=7, fig.width=9}
plot_dates_object <- CB_Data %>%
  # use the variables of the dates and make a longer dataset
  # In the pivot_longer() command we select the columns which 
  # we want to expand in long format and transform the dataset 
  pivot_longer(
    # all columns starting with "DateOf..." will be taken 
    cols=starts_with("DateOf"), 
    # the names of the variables will be placed in a single column called "Indicator"
    names_to = "Indicator", 
    # the values (which are dates in this case) will be placed in a 
    # column called "Date
    values_to = "Date") %>%
  # group by date and indicator (type of date, i.e., onset, diagnosis or notification) 
  group_by(Indicator, Date) %>% 
  summarise(n=n(), .groups="drop") %>% 
  # we add a filter to remove any rows where a date value might not exist
  filter(!is.na(Date))

# then we pass the object to the ggplot()
plot_dates <- plot_dates_object %>% 
  ggplot() +
  geom_line(aes(x = Date, y = n, color=Indicator)) +
  scale_x_date(breaks = "2 weeks")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        axis.text.x = element_text(angle=90),
        legend.title = element_blank()) +
  labs(title="E-pox cases reported in 2022, by date of onset, diagnosis and notification.")

plot_dates
```

Then, we try to plot by week and there is some improvement in the visualization.

```{r E-pox-Rpractical-11, fig.height=7, fig.width=9}
plot_date_w<- CB_Data %>%
  # use the variables of the dates and make a longer dataset
  pivot_longer(cols=starts_with("DateOf"), names_to = "Indicator", values_to = "Date") %>%
  mutate(Date=floor_date(Date, unit="week", week_start="Monday")) %>% 
  # group by date and indicator (type of date, i.e., onset, diagnosis or notification) 
  group_by(Indicator, Date) %>% 
  summarise(n=n(), .groups="drop") %>% 
  filter(!is.na(Date)) %>% 
  ggplot() +
  geom_line(aes(x = Date, y = n, color=Indicator)) +
  scale_x_date(breaks = "2 weeks")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        axis.text.x = element_text(angle=90),
        legend.title = element_blank()) +
  labs(title="E-pox cases reported in 2022, by date of onset, diagnosis and notification.")

plot_date_w
```


We can create the same graph by country (selecting only 2 countries).

```{r E-pox-Rpractical-12, fig.height=15, fig.width=9}

plot_date_w_country<- CB_Data %>%
  filter(Country=="CountryA"|Country=="CountryE") %>% 
  # use the variables of the dates and make a longer dataset
  pivot_longer(cols=starts_with("DateOf"), names_to = "Indicator", values_to = "Date") %>%
  mutate(Date=floor_date(Date, unit="week", week_start="Monday")) %>% 
  # group by date and indicator (type of date, i.e., onset, diagnosis or notification) 
  group_by(Country, Indicator, Date) %>% 
  summarise(n=n(), .groups="drop") %>% 
  filter(!is.na(Date)) %>% 
  ggplot() +
  geom_line(aes(x = Date, y = n, color=Indicator)) +
  # in this line we indicate that a graph should be created 
  # by country and that the axis should have different scales which
  # are adjusted by the range of values in each country
  facet_wrap(~Country, ncol = 1, scales = "free")+
  scale_x_date(breaks = "2 weeks")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        axis.text.x = element_text(angle=90),
        legend.title = element_blank()) +
  labs(title="E-pox cases reported in 2022, by date of onset, diagnosis and notification.")

plot_date_w_country
```

#### Graph on cases reported by day and country

```{r E-pox-Rpractical-13,  fig.height=5, fig.width=9}

hm_CB_Data<-CB_Data %>% 
  distinct(DateOfNotification, Country) %>% 
  mutate(n=1) %>%  
  ungroup() %>% 
  ggplot(.) +
  geom_tile(aes(x=DateOfNotification, y=as.factor(Country), fill = as.factor(Country)), alpha=0.7) +
  geom_hline(yintercept = seq(.5, nlevels(as.factor(CB_Data$Country)), 1), linewidth = .2) +
  geom_vline(xintercept = max(CB_Data$DateOfNotification, na.rm=T)-as.difftime(tim = 21, units = "days"), linetype=2) +
  geom_vline(xintercept = max(CB_Data$DateOfNotification, na.rm=T)-as.difftime(tim = 30, units = "days"), linetype=3) +
  geom_vline(xintercept = max(CB_Data$DateOfNotification, na.rm=T), linetype=1) +
  scale_fill_hue(l=4) +
  scale_x_date(date_breaks = "1 month", date_labels = "%B") +
  theme_minimal() +
  theme(legend.position = "none", 
        axis.text = element_text(size=9),
        axis.title = element_blank()) +
  labs(caption="Vertical lines: 21 and 30 days before last case notification.",
       title="Date of reporting of E-pox cases by country.")

hm_CB_Data

```


### Demographics

```{r E-pox-Rpractical-14, echo=TRUE, fig.height=7, fig.width=9}
CB_Data <- CB_Data %>% 
  mutate(Age_cat=cut(Age, 
                     breaks = c(0, 14, 17, 30, 40, 50, 60, 999), 
                     include.lowest = T))

CB_demo <- CB_Data %>%
  # Select demographics and other columns
  select("Age_cat") %>%
 # Create summary table (attention to handling of missing values)
  gtsummary::tbl_summary(missing = "ifany")

CB_demo
```

### Clinical description

```{r E-pox-Rpractical-15, echo=TRUE, fig.height=7, fig.width=9}

clin_desc_table <- CB_Data %>%
  # Select demographics and other columns
  select("ClinicalSymptoms") %>%
 ## Create summary table (attention to handling of missing values)
  gtsummary::tbl_summary(missing = "ifany") 

clin_desc_table

clinical_desc <- CB_Data %>%  
  filter(ClinicalSymptoms!="") %>% 
  # Group by symptoms
  group_by(ClinicalSymptoms) %>% 
  # Count for each symptom:
  summarise(count = n(), .groups="drop") %>% 
  mutate(prop=(count/sum(count))*100) %>% 
  # Create plot:
  ggplot(aes(
    # ordering the symptoms by count
    x = reorder(ClinicalSymptoms, desc(prop), decreasing = TRUE), 
    y = prop)) +
  # Display bars as proportions
  geom_bar(stat = "identity") +
  #x axis label
  xlab("Symptoms") +
  #y axis label
  ylab("% E-pox cases") +
  # flip the axis
  coord_flip() +
  theme_minimal()

# Print plot:
clinical_desc

```


### Outcome, HIV status

```{r E-pox-Rpractical-16, echo=TRUE}
Outcome_HIV <- CB_Data %>%
  # Select columns
  select("Outcome", "HIVStatus") %>%
  mutate(Outcome=case_when(Outcome=="UNK"~"Unknown",
                           Outcome=="A"~"Alive",
                           .default = Outcome))  %>%
  mutate(HIVStatus=case_when(HIVStatus=="POS"~"Positive",
                           HIVStatus=="NEG"~"Negative",
                           .default = HIVStatus)) %>% 
 # Create summary table (attention to handling of missing values)
  gtsummary::tbl_summary(by = Outcome) %>%
  gtsummary::add_overall() 
Outcome_HIV

```

### Sexual orientation

```{r E-pox-Rpractical-17, echo=TRUE}
sexorient <- CB_Data %>%
  filter(!is.na(SexualOrientation)) %>% 
  # Select columns
  select("SexualOrientation") %>%
  gtsummary::tbl_summary() 

sexorient <- CB_Data %>%
  filter(!is.na(SexualOrientation)) %>% 
  # Select columns
  select("SexualOrientation") %>%
  mutate(SexualOrientation=case_when(SexualOrientation=="BISEXUAL"~"Bisexual",
                                     SexualOrientation=="HETERO"~"Heterosexual",
                                     SexualOrientation=="MSM"~"MSM/homo or bisexual male",
                                     SexualOrientation=="UNK"~"Unknown or undetermined",
                                     .default=SexualOrientation)) %>% 
  gtsummary::tbl_summary() 


sexorient


# Optional - Creating the age and gender pyramid (first filtering out non M or F rows)
agegender_pyramid <- CB_Data %>%
  filter(Gender=="Female"|Gender=="Male" & !is.na(Age_cat)) %>% 
  # Create age gender pyramid
  apyramid::age_pyramid(
  # Specify column containing age groups
  age_group = "Age_cat",
  # Splitting by Gender
  split_by = "Gender", 
  # Don't show midpoint on the graph:
  show_midpoint = FALSE) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x="")


agegender_pyramid



```


## (Optional) Exploring trends and progress

In this part of the analysis we go beyond describing the basic information bue we are also checking how the trends of the reported cases might have changed. Therefore, we work with data reported before the last 3 weeks in which the data collected might be incomplete.

```{r E-pox-Rpractical-18}
week_xx <- paste0(format(as.Date(
  max(CB_Data$WeekDate) - as.difftime(tim = 3, units = "weeks")
), format = "%d/%m"),
"-",
format(as.Date(
  max(CB_Data$WeekDate) - as.difftime(tim = 15, units = "days")
),
format = "%d/%m"))
week_yy <-
  paste0(format(as.Date(
    max(CB_Data$WeekDate) - as.difftime(tim = 4, units = "weeks")
  ), format = "%d/%m"),
  "-",
  format(as.Date(
    max(CB_Data$WeekDate) - as.difftime(tim = 22, units = "days")
  ), format = "%d/%m"))

WeekCB_Data <- CB_Data %>%
  group_by(WeekDate) %>%
  summarise(n = n(), .groups="drop")

ltw <- WeekCB_Data %>%
  filter(
    WeekDate < max(CB_Data$WeekDate) - as.difftime(tim = 2, units = "weeks") &
      WeekDate >= max(CB_Data$WeekDate) - as.difftime(tim = 4, units = "weeks")
  ) %>%
  mutate(w_n = case_when(
    WeekDate == as.Date(max(CB_Data$WeekDate) - as.difftime(tim = 3, units = "weeks")) ~
      "current",
    WeekDate == as.Date(max(CB_Data$WeekDate) - as.difftime(tim = 4, units = "weeks")) ~
      "previous"
  ))


week_max <-
  paste0(format(as.Date(filter(
    WeekCB_Data, n == max(WeekCB_Data$n)
  )$WeekDate), format = "%d/%m"),
  "-",
  format(as.Date(
    filter(WeekCB_Data, n == max(WeekCB_Data$n))$WeekDate + as.difftime(tim = 6, units = "days")
  ), format = "%d/%m"))


# Days since last case

days_since_last_case <- CB_Data %>%
  group_by(Country) %>%
  summarise(days_last_report = as.integer(difftime(
    time1 = max(CB_Data$DateOfNotification, na.rm = T),
    time2 = max(DateOfNotification),
    units = "days"
  )), .groups="drop") %>%
  ungroup() %>%
  mutate(
    category = case_when(
      days_last_report <= 21 ~ "21 or less",
      days_last_report <= 30 ~ "22-30 days",
      days_last_report > 30 ~ "More than 30 days"
    )
  )


summary_days_since_last <- days_since_last_case %>%
  group_by(category) %>%
  summarise(n_countries = n(), .groups="drop") %>%
  adorn_totals()

```


There was a `r round(((filter(ltw, w_n=="current")$n-filter(ltw, w_n=="previous")$n)/filter(ltw, w_n=="previous")$n)*100, digits=1)`% change in the number of cases between `r week_xx` (`r filter(ltw, w_n=="current")$WeekDate`, n=`r filter(ltw, w_n=="current")$n`) and `r week_yy` (`r filter(ltw, w_n=="previous")$WeekDate`, n=`r filter(ltw, w_n=="previous")$n`).

The highest number of cases since the beginning of the outbreak was reported on `r week_max` (`r filter(WeekCB_Data, n==max(WeekCB_Data$n))$WeekDate`, n=`r filter(WeekCB_Data, n==max(WeekCB_Data$n))$n`). 

+ Change in the number of cases since the week with the max number of cases: `r round(((filter(ltw, w_n=="current")$n-max(WeekCB_Data$n))/max(WeekCB_Data$n))*100, digits=1)`% (difference: `r (filter(ltw, w_n=="current")$n-max(WeekCB_Data$n))` cases)


`r kable(summary_days_since_last, caption=paste0("Number of countries and days since last case and until ", max(CB_Data$DateOfNotification)))`

`r kable(days_since_last_case, caption=paste0("Number of countries and days since last case and until ", max(CB_Data$DateOfNotification)," - complete table"))`



# (Optional). Combine case-based and aggregated data and generate a summary of the information


```{r E-pox-Rpractical-19}
# Get the weekly data and prepare them

# --- for the aggregated ones, add the new cases from the cumulative
Agg_Data_weekly <- Agg_Data %>% 
  rename(Cases_total="Cases") %>% 
  mutate(Source="OpenData") %>% 
  group_by(Country)  %>% 
  arrange(Country, DateRep) %>%
  # after arranging/sorting we have to subtract the previous cumulative 
  # cases from the "current"
  # since for the first value this cannot happen we use the starting cumulative number
  mutate(Cases=Cases_total-lag(Cases_total)) %>% 
  mutate(Cases=case_when(is.na(Cases)~Cases_total,
                             .default = Cases))


# --- for the case-based data we need to summarise and then add the cumulative number
CB_Data_weekly <- CB_Data %>% 
  group_by(Country, WeekDate) %>% 
  arrange(WeekDate) %>% 
  summarise(Cases=n(), .groups="drop") %>% 
  ungroup() %>% 
  group_by(Country) %>% 
  mutate(Cases_total=cumsum(Cases),
         Source="ReportedCases") 


# Check the cumulative cases by week in case-based and aggregated data
# for the most recent week
Agg_Data_total <- Agg_Data_weekly %>% 
  group_by(Country) %>% 
  filter(WeekDate==max(WeekDate)) %>% 
  select(c("Country", "Cases_total", "Cases", "Source"))


CB_Data_total <- CB_Data_weekly %>% 
  group_by(Country) %>% 
  filter(WeekDate==max(WeekDate)) %>%
  select(c("Country", "Cases_total", "Cases", "Source"))

# using the combined dataset we make it wide and check the difference between 
# the case-based and the aggregated data
combined_df <- bind_rows(Agg_Data_total, CB_Data_total)

combined_df_wide <- combined_df %>% 
  select(-Cases) %>% 
  pivot_wider(id_cols = "Country", 
              names_from = "Source", values_from = "Cases_total", values_fill = 0) %>% 
  mutate(difference=OpenData-ReportedCases)

# we create a plot with the data by source and country to visualise the difference
comparison_plot = ggplot() +
    geom_col(data = combined_df, aes(x = Source, y = Cases_total, fill = Source)) +
    facet_wrap("Country", scales = "free_y")

comparison_plot

# select for which countries we will use the case-based data
cb_countries <- combined_df_wide %>% 
  filter(OpenData <= ReportedCases) %>% 
  pull(Country)


# Make a list of the countries for which case-based data were used 
list_CB_countries <- paste(cb_countries, collapse = ", ")

# from the case based data we select only the countries we want and add them to the aggregated
cb_selected <- CB_Data_weekly %>% 
  filter(Country %in% cb_countries) %>% 
  select(Country, WeekDate, Cases_total, Cases, Source)

# From the aggregated data remove the countries we do not need 
Agg_selected <- Agg_Data_weekly %>% 
  filter(!Country %in% cb_countries) %>% 
  select(Country, WeekDate, Cases_total, Cases, Source)

# Create the final dataset with the selected countries from each source
sum_db <- bind_rows(cb_selected, Agg_selected)

```

# (Optional) Summary of the outbreak based on the both the case-based and aggregated data

```{r E-pox-Rpractical-20}

since_date<- glue("{day(max(sum_db$WeekDate)-weeks(3))} {month(max(sum_db$WeekDate)-weeks(3), label = T, abbr = F)} {year(max(sum_db$WeekDate)-weeks(3))}")
as_of_date<- glue("{day(max(sum_db$WeekDate))} {month(max(sum_db$WeekDate), label = T, abbr = F)} {year(max(sum_db$WeekDate))}")

# In last 4 weeks - since we use the weekly data
since_last_update <- sum_db %>% 
  # taking the data of the last 4 weeks
  filter(WeekDate>=max(sum_db$WeekDate)-weeks(3)) %>% 
  group_by(Country) %>% 
  summarise(n=sum(Cases), .groups = "drop") %>% 
  arrange(desc(n))
  
since_last_update_total <- sum(since_last_update$n)
  
since_last_update_sentence <- paste0(paste0(since_last_update$Country, 
                                            " (", since_last_update$n, ")"),
                                     collapse = ", ") %>%
  stringi::stri_replace_last(fixed = ",", replacement = " and")
```

##  E-pox situation update, as of the week of `r as_of_date`

### Update

Since the last update on the week of `r since_date`, and as of the week of `r as_of_date`, `r since_last_update_total` E-pox cases have been reported from `r nrow(since_last_update)` countries: `r since_last_update_sentence`.


```{r E-pox-Rpractical-21, echo=FALSE}

# Estimating the total number of cases from the final dataset
totals <- sum_db %>% 
  group_by(Country) %>% 
  summarise(Total_cases=sum(Cases), .groups="drop") 

# Making a list of the countries and their total number of cases to be used in the summary
totals_sentence <- paste0(paste0(totals$Country, 
                                            " (", totals$Total_cases, ")"),
                                     collapse = ", ") %>%
  stringi::stri_replace_last(fixed = ",", replacement = " and")

```

### Summary

Since the start of the outbreak and as of `r as_of_date`, `r sum(sum_db$Cases)` E-pox cases have been reported from `r n_distinct(sum_db$Country)` countries: `r totals_sentence`.


### Epicurve

```{r E-pox-Rpractical-22, echo=FALSE}
epicurve_db <- sum_db %>% 
  group_by(WeekDate) %>% 
  summarise(Cases=sum(Cases)) %>% 
  ungroup() 

epicurve_all <- epicurve_db %>% 
  mutate(Completeness=case_when(WeekDate<=max(sum_db$WeekDate)-days(21)~"Complete",
                                .default="Incomplete")) %>% 
  ggplot() + 
  # Add geom_bar:
  geom_col(aes(x = WeekDate, y=Cases, fill=Completeness)) + 
  scale_x_date(date_breaks = "4 week") +
  scale_fill_manual(values = c("#65B32E","#9CC65A")) +
  # Remove unnecessary grid lines:
  theme_bw()+
  theme(legend.position = "bottom")+
  # Update x and y axis labels:
  labs(x = "Reporting Week", 
       y = "Number of cases",
       title="E-pox cases") 

epicurve_all

```


Disclaimer: The countries for which case-based data were used are the following: `r list_CB_countries`. For all other countries, data were collected from official, open sources.
