---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Case-studies Open Repository {.unnumbered}

```{r, echo=F}
pacman::p_load(
  rio,
  here,
  janitor,
  DT,
  tidyverse
)
```

**Objective**: This open repository of case studies can be used as a guide on how to implement different epidemiological methods in different situations using a variety of tools

# How to use this repository {.unnumbered}

# Acknowledgements {.unnumbered}

## Authors {.unnumbered}


# List of Case-studies {.unnumbered}


This is another kind of internal link that makes use of a tag: [Fulton @sec-fulton-en]

This is a static table to test links:


```{r, echo=F}
table_cases <- tribble(
  ~Name, ~Link, ~Geography, ~Scale, ~Tool, ~Language, ~Topics,
  "Fulton COVID-19", 'pages/fulton-en.html', "USA", "Local", "R", "English", "Outbreak reports; GIS; epidemic curves", 
  "Oswego ES", "pages/oswego-es.html", "USA", "Local", "R", "Spanish", "Outbreak response"
)

```

Filterable version of table (but will need to insert weblinks, not references as they do not seem to work)

```{r, echo = F, eval = T}



# Render the table with DT, making the 'Link' column clickable with custom text
datatable(table_cases, escape = FALSE, filter = "top", options = list(
  columnDefs = list(
    list(targets = 2, render = JS(
      "function(data, type, row, meta) {
        if(type === 'display') {
          var linkText = 'Link'; // Use the 'Name' column as the display text
          data = '<a href=\"' + data + '\">' + linkText + '</a>';
        }
        return data;
      }"
    ))
  )
), callback = JS("table.page.len(5).draw();"))

```




